\sysfull

### 2018-08-06 paper/asplos19/implementation.tex

We implemented \sys's programming and execution model as a runtime library and API that a programmer can use to make a plain C program intermittence-safe.
%
\subsection{Application Programming Interface}
\label{sec:coala_api}
%
Coala's API adds only eight syntactic constructs to a C-based language, summarized
in Table~\ref{table:coala_api}.
%
\begin{table}
\centering
\begin{tabular}{| r | p{0.67\columnwidth} |}
	\hline
	{Method} & {Arguments} \\
	\hline\hline
	\texttt{INIT}($t$) & $t \in \mathbf{T}$: scheduled task on first boot \\
	\hline
	\texttt{RUN}() & --- \\
	\hline
	\texttt{TASK}($t$, $w_t$) & $t \in \mathbf{T}$: task name, $w_t$: weight of task $t$ \\
	\hline
	\texttt{NEXT\_TASK}($t$) & $t \in \mathbf{T}$ : task to be run next \\
	\hline
	\texttt{PV}($p$, $v$ [, $s$]) & $p$: variable type, $v \in \mathbf{V}$: new protected variable name, $s$: array size \\
	\hline
	$u$ := \texttt{RP}($v$) & $v \in \mathbf{V}$: protected variable to be read, $u$: destination operand \\
	\hline	
	\texttt{WP}($v$) := $u$ &  $v \in \mathbf{V}$: protected variable to be written, $u$: source operand \\
	\hline
	\texttt{SM}($p$, $m$ [, $s$]) & $p$: \texttt{struct} member's type, $m$: \texttt{struct} member's name, $s$: array size \\
	\hline
\end{tabular}
\caption{API summary; $\mathbf{T}$: set of all tasks, $\mathbf{V}$: set of all protected variables, $[, s]$: optional argument.}
\label{table:coala_api}
\end{table}
%
\noindent\textbf{New Tasks.} The \texttt{TASK} annotation on a function
declaration statically allocates a non-volatile constant variable holding a
task's weight and declares that the function is a task.\\
%
\noindent\textbf{Task Transitions.} A call to \texttt{NEXT\_TASK} marks the task to be executed at the 
end of the current one task and it can be invoked along
any control path to dynamically determine the next task.\\
%
\noindent\textbf{Protected Variables.} The \texttt{PV} annotation on a
variable statically allocates a protected non-volatile variable. The
variable must then be accessed with the \texttt{RP} and \texttt{WP} API methods at
runtime to ensure correct operation.\\
%
\noindent\textbf{Initialization.} The behavior of the API method
\texttt{INIT} is very similar to \texttt{NEXT\_TASK}, with the
addition of performing preliminary kernel initializations, including hardware setup.\\
%
\noindent\textbf{Execution.} The programmer passes control to \sys's task scheduler 
by calling \texttt{RUN} after device initialization. 
%
First the scheduler updates the coalescing target using the coalescing
strategy.  Second, the scheduler resumes a commit, if one is in progress. Third,
the scheduler clears the list of dirty shadow pages. The scheduler then 
sets the program counter to the next task to run, which \sys tracks in  
non-volatile memory.  
%
Before executing the task, \sys checks whether there is a partially committed
task to resume, which requires \sys to restore volatile state, including the program
counter.  
%
If there is no in-progress, partially committed task, \sys starts executing and
coalescing tasks.

@@

As described in Section~\ref{sec:memory_virtulaization}, \sys's memory
virtualization scheme uses three buffers: a volatile \texttt{working} buffer, a
non-volatile \texttt{\underline{private}}, and a non-volatile
\texttt{\underline{shadow}} buffer. 
%
The main memory \texttt{\underline{private}} holds the last committed copy of all 
pages. 
%
After reboot, the \texttt{working} buffer is empty. Execution fills 
the \texttt{working} buffer with privatized pages as the program accesses them. 
%
The \texttt{\underline{shadow}} buffer serves as an intermediate,
non-volatile backing \texttt{\underline{private}} for dirty pages evicted from the space-constrained
\texttt{working} buffer.

To access a protected variable with either \texttt{RP}
or \texttt{WP}, \sys searches for the variable in its \texttt{working} buffer first.
 If the page is not, the VMM fetches it from non-volatile memory. 
%
\sys uses address-based page tagging to make finding a variable efficient.  The
upper bits of a variable's memory address identify its page, and the lower bits
denote the variable's offset in its page. The total number of pages in memory,
$p$, determines the number of tag bits, which is $\log P$.

Page tagging imposes a data alignment requirement.
Page size $S$ must be a power of two. 
%
Pages must be aligned to an $S$-byte boundary for efficient memory access. 
%
We experimented with 32-, 64-, 128- and 256\,B pages, an 8\,KB non-volatile \texttt{\underline{shadow}} buffer and
an 8\,KB non-volatile \texttt{\underline{private}}, and a \texttt{working} buffer
of 1\,KB.
%
\subsection{Coalescing and Commit}
%
The coalescing algorithm starts with a random guess about the size of the coalescing task, in our implementation the target is initialized to four. It then refines its guess based on the number of the static-tasks being successfully executed.
%
When a coalesced task completes, \sys save all the modifications in its \texttt{\underline{private}} using DMA. In order to ensure the atomicity of the commit operation all the modified pages are copied to the \texttt{\underline{shadow}} buffer first. The \sys set a flag to indicates an on-going \emph{atomic commit} to update the private pages. For efficiency reasons \sys uses an indirection table to point to the private consistent pages and eliminate the need for moving pages for the second time.
%
\subsection{Task Downscaling}
%
\sys's task downscaling preserves forward progress by partially committing an
ongoing, non-terminating task. 
%
Partial commit relies on the existence of a hardware timer, which is a very
common feature in a typical MCU and consumes very little energy (i.e.,
$\approx$3\,$\mu$A/MHz~\cite{msp430datasheet}).

 \textbf{Partial Commit and Task Atomicity.} Some applications prevent
downscaling a task because of a need for task atomicity.  For example, sampling
an analog signal requires consecutive samples at a known interval, or the
digitally sampled signal is meaningless.  In such a case, the programmer can
disable partial commit for a task or a span of code, marking  the code with a
pair of \texttt{DISABLE\_PE} and \texttt{ENABLE\_PE} annotations.
These annotations respectively halt and resume the partial execution timer.

>>>>

We implemented \sys's programming and execution model as a runtime library and API that a programmer can use to make a plain C program intermittency-safe.
%
\subsection{Application Programming Interface}
\label{sec:coala_api}
%
Coala's API adds only a few syntactic constructs to a C-based language, summarized
in Table~\ref{table:coala_api}.
%
\begin{table}
\footnotesize
\centering
\begin{tabular}{| r | p{0.7\columnwidth} |}
	\hline
	{Method} & {Arguments} \\
	\hline\hline
	\texttt{INIT}($t$) & $t \in \mathbf{T}$: scheduled task on first boot \\
	\hline
	\texttt{RUN}() & --- \\
	\hline
	\texttt{TASK}($t$, $w_t$) & $t \in \mathbf{T}$: task name, $w_t$: weight of task $t$ \\
	\hline
	\texttt{NEXT\_TASK}($t$) & $t \in \mathbf{T}$ : task to be run next \\
	\hline
	\texttt{PV}($p$, $v$ [, $s$]) & $p$: type, $v \in \mathbf{V}$: name, $s$: array size \\
	\hline
	$u$ := \texttt{RP}($v$) & $v \in \mathbf{V}$: protected variable to read, $u$: dest.\ operand \\
	\hline	
	\texttt{WP}($v$) := $u$ &  $v \in \mathbf{V}$: protected variable to write, $u$: source operand \\
	\hline
	\texttt{SM}($p$, $m$ [, $s$]) & $p$: type, $m$: name, $s$: array size \\
	\hline
	\texttt{DISABLE\_PC}() & --- \\
	\hline
	\texttt{ENABLE\_PC}() & --- \\
	\hline
\end{tabular}
\caption{API summary; $\mathbf{T}$: set of all tasks, $\mathbf{V}$: set of all protected variables, $[, s]$: optional argument.}
\label{table:coala_api}
\end{table}

\paragraph{New Tasks.} The \texttt{TASK} annotation on a function
declaration statically allocates a non-volatile constant variable holding a
task's weight and declares that the function is a task.

\paragraph{Task Transitions.} \texttt{NEXT\_TASK} marks the task to be executed at the 
end of the current one task and it can be invoked along
any control path to dynamically determine the next task.

\paragraph{Protected Variables.} The \texttt{PV} annotation on a
variable statically allocates a protected non-volatile variable. The
variable must then be accessed with the \texttt{RP} and \texttt{WP} API methods at
run-time to ensure correct operation.

\paragraph{Initialization.} The behavior of the API method
\texttt{INIT} is very similar to \texttt{NEXT\_TASK}, with the
addition of performing preliminary kernel initializations, including hardware setup.

\paragraph{Execution.} The programmer passes control to \sys's task scheduler 
by calling \texttt{RUN} after device initialization. 
%
First the scheduler updates the coalescing target using the coalescing
strategy.  Second, the scheduler resumes a commit, if one is in progress. Third,
the scheduler clears the list of dirty shadow pages. The scheduler then 
sets the program counter to the next task to run, which \sys tracks in  
non-volatile memory.  
%
Before executing the task, \sys checks whether there is a partially committed
task to resume, which requires \sys to restore volatile state, including the program
counter.  
%
If there is no in-progress, partially committed task, \sys starts executing and
coalescing tasks.

%
\subsection{Task Coalescing}
%
% The coalescing algorithm starts with a random guess about the size of the coalescing task, in our implementation the target is initialized to four. It then refines its guess based on the number of the static-tasks being successfully executed.
%
\paragraph{Parameters.}
Equations (\ref{eq:eo}) and (\ref{eq:eg}) parametrize the behavior of the coalescing strategies in terms of $x$, $\rho$ and $\gamma$.
We experimented with a range of values, and then used the ones that yielded the best performance: $x = 1$ (for EO) and $\rho = \gamma = 1$ (for EG/WEG).

\paragraph{Weights for WEG.}
The effectiveness of the WEG hinges on correctly identifying the weight of each task, which WEG assumes is {\em statically} available. Profiling the time and energy cost of tasks in a program is a difficult, orthogonal problem~\cite{cleancut_2018,baghsorkhi_cgo_2018}. WEG could use the result of an arbitrarily sophisticated profiling procedure. To produce a concrete result in this paper, we give WEG access to a simple profile of task run time collected offline using a single fixed input. WEG stores the profile in a lookup table that maps a task's identifier to its weight, making the information available to \sys's scheduler at run time.

%
\subsection{Task Downscaling}
%
% \sys's task downscaling preserves forward progress by partially committing an
% ongoing, non-terminating task. 
%
\paragraph{Timer Strategy.}
After identifying a task as likely non-terminating, \sys must decide when during the task's execution to partially commit the task's state. \sys sets a timer at the start of the likely non-terminating task, initializing it to a very large value (e.g., an estimate of the maximum execution time using the device's energy buffer). If the timer expires before power fails, \sys partially commits, retaining  the timer value to use for future partial commits. If the timer does not expire before power fails, \sys halves the timer and tries again. The exponential decrease of the timer value converges rapidly to a usable one.
%
Partial commit relies on the existence of a hardware timer, which is a very
common feature in a typical MCU and consumes very little energy (i.e.,
$\approx$3\,$\mu$A/MHz~\cite{msp430datasheet}).

 \paragraph{Partial Commit and Task Atomicity.}
 Some applications may prevent
downscaling a task because of a need for task atomicity.  For example, sampling
an analog signal requires consecutive samples at a known interval, or the
digitally sampled signal is meaningless.  In such a case, the programmer can
disable partial commit for a task or a span of code, marking  the code with a
pair of \texttt{DISABLE\_PC} and \texttt{ENABLE\_PC} annotations.
These annotations respectively halt and resume the partial commit timer.

@@

% As described in Section~\ref{sec:memory_virtulaization}, \sys's memory
% virtualization scheme uses three buffers: a volatile \texttt{working} buffer, a
% non-volatile \texttt{\underline{private}}, and a non-volatile
% \texttt{\underline{shadow}} buffer. 
%
% The main memory \texttt{\underline{private}} holds the last committed copy of all 
% pages. 
%
% After reboot, the \texttt{working} buffer is empty. Execution fills 
% the \texttt{working} buffer with privatized pages as the program accesses them. 
%
% The \texttt{\underline{shadow}} buffer serves as an intermediate,
% non-volatile backing \texttt{\underline{private}} for dirty pages evicted from the space-constrained
% \texttt{working} buffer.

% To access a protected variable with either \texttt{RP}
% or \texttt{WP}, \sys searches for the variable in its \texttt{working} buffer first.
 % If the page is not, the VMM fetches it from non-volatile memory. 
%
\paragraph{Efficiency.}
\sys uses address-based page tagging to make finding a variable efficient.  The
upper bits of a variable's memory address identify its page, and the lower bits
denote the variable's offset in its page. The total number of pages in memory,
$P$, determines the number of tag bits, which is $\log_2 P$.
%
Furthermore, the VMM moves pages of data efficiently using hardware accelerated DMA support.

\paragraph{Alignment.}
Page tagging imposes a data alignment requirement.
The page size $S$ must be a power of two. 
%
Pages must be aligned to an $S$-byte boundary for efficient memory access. 
%
To preserve alignment, when typedef'ing a C structure for protected variables, the programmer has to use the API method \texttt{SM} on all members of the structure (only when defining the structure).
%
We experimented with 32-, 64-, 128- and 256\,B pages, an 8\,KB non-volatile \texttt{\underline{shadow}} buffer and
an 8\,KB non-volatile \texttt{\underline{private}} buffer, and a \texttt{working} buffer
of 1\,KB.

\paragraph{Page Eviction.}
When a page in \texttt{working} has to be swapped out to make room for a newly requested one, an eviction policy has to be chosen.
While we opted for the simplicity of FIFO, any other replacement policy, such as Least Recently Used (LRU), could work in its place.

### 2018-08-06 paper/asplos19/partialcommit.tex
revise partial commit
Downscaling section revised

\sys uses {\em task downscaling} to make progress through a task that is too
large to complete using the energy that the device can buffer.  Task
downscaling executes part of the long task and interrupts its execution with a
{\em partial commit}, after which the long task continues. If power fails, the
partial commit preserves the progress through the task, and execution resumes
from the point of the partial commit, rather than from the start of the task.
Eventually, after some number of partial commits and power failures, the task
will complete and execution will proceed to a task with an energy demand
accommodated by the device's energy buffer. Task downscaling violates task
atomicity in a non-terminating execution, favoring continued progress over
atomicity. If a program requires a task's atomicity over its progress, 
the programmer can disable task downscaling for that task.

The key design issue for task downscaling is deciding when to partially commit.
\sys only applies task downscaling to a task that is likely to be
non-terminating.  A task is likely to be non-terminating if it fails to execute
completely twice consecutively. The second incomplete execution executes after
a power failure, when the device will have fully recharged its energy buffer.
If a task cannot complete when executing with a full energy buffer, the task is
non-terminating.

After identifying a task as likely non-terminating, \sys must decide when
during the task's execution to partially commit the task's state.
%
\sys sets a timer at the start of the likely non-terminating task, initializing
it to a very large value (e.g., an estimate of the maximum execution time using
the device's energy buffer).
%
If the timer expires before power fails, \sys partially commits, retaining 
the timer value to use for future partial commits. 
%
If the timer does not expire before power fails, \sys halves the timer and
tries again.
%
\sys's exponential decrease of the timer value converges rapidly to a useable
timer value.

%Putting these in comments to make space look right
%\textcolor{red}{More variables describing the partial commit algorithm: we do not explore as many options for partial commit as we do for coalescing}

%\todo{Proof of the maximum granularity of task downscaling - show it is limited by the clock}{Carlo}

%\subsubsection{Why Task Downsizing is not \sys's Default Behavior}
%I do not understand what this section is about so I am removing it.  If someone wants to take another crack at re-writing it feel free.
%Task downsizing is the core novelty of \sys when it comes to code portability and re-usability. However, it has limitations, which force  it to be a non-default \sys behavior. 
%
%\textbf{Static deployment of partial commit stubs.} This approach reduces programmer effort, although the programmer still needs to apply \sys's API and distribute trigger points for partial commits. On the other hand, \sys' downsizing introduces an additional overhead by saving the stack and the register file each time it commits the pages. 
%
%\textbf{Dynamic deployment of partial commit stubs.} \sys downsizing does not rely on any information from the programmer to trigger the commit process. However, \sys utilizes a trial and error approach to optimize the commit rate over the entire program execution time. Therefore, it suffers from a large cumulative re-execution penalty.

>>>>

\sys uses {\em task downscaling} to make progress through a task that is too large to complete using the energy that the device can buffer. Task downscaling executes part of the long task and interrupts its execution with a {\em partial commit}, after which the long task continues. If power fails, the partial commit preserves the progress through the task, and execution resumes from the point of the partial commit, rather than from the start of the task. Eventually, after some number of partial commits and power failures, the task will complete and execution will proceed. Task downscaling violates task atomicity in a non-terminating execution, favoring continued progress over atomicity. If a program requires a task's atomicity to be preserved, they can disable task downscaling for that task.

The key design issue for task downscaling is deciding when to partially commit. \sys only applies task downscaling to a task that is likely to be non-terminating.  A task is likely to be non-terminating if it fails to execute completely twice consecutively. The second incomplete execution executes after a power failure, when the device will have fully recharged its energy buffer. If a task cannot complete when executing with a full energy buffer, the task is marked as non-terminating.

% After identifying a task as likely non-terminating, \sys must decide when during the task's execution to partially commit the task's state. \sys sets a timer at the start of the likely non-terminating task, initializing it to a very large value (e.g., an estimate of the maximum execution time using the device's energy buffer). If the timer expires before power fails, \sys partially commits, retaining  the timer value to use for future partial commits. If the timer does not expire before power fails, \sys halves the timer and tries again. The exponential decrease of the timer value converges rapidly to a usable one.
